---
title: "MPO Reference Range"
output:
  html_document: 
      toc: true
      toc_float: 
        collapsed: false
      code_folding: hide
  html_notebook: default
---

<!-- Custom CSS Styles -->
<style type="text/css">
.table {

    width: auto;
    margin-left: 1in;

}
</style>
```{r functions for stu, echo=FALSE, message=F}
percentile.table <- function(df, p=c(.01,.05,.10,.25,.33,.50,.67,.75,.90,.95,.99)) {
	df = as.data.frame(df)
	as.data.frame(sapply(colnames(df), function(cn) {
		quantile(as.numeric(as.character(df[[cn]])), p, na.rm=T)
	}))
}

add.test.col <- function(mdat, test.data, abbr, result.code, name=NA, units=NA, type=NA, reported=NA, trim.na=FALSE) {
	if (length(test.data) != nrow(mdat))
		stop("test.data length doesn't equal rows in mdat")
  if (missing(abbr) | missing(result.code))
    stop("Must provide abbr and result.code")
  test.name = sprintf("%s.code%s", abbr, result.code)
  if (test.name %in% names(mdat))
    stop("That test.name is already in the mdat")
	if (!any(grepl("[.]code", test.name)))
	  warning("test name doesn't have the .code substring, so it won't be treated as a test by get.tests(). Consider renaming")
	test.data[test.data=="None"] = NA
	
	local.col.meta = get.col.meta(mdat)
	local.col.meta[ test.name, "Name"] = name
	local.col.meta[ test.name, "UnitsofMeasure"] = units
	local.col.meta[ test.name, "TestType"] = type
	local.col.meta[ test.name, "Reported"] = reported
	local.col.meta[ test.name, "Abbrev"] = abbr
	local.col.meta[ test.name, "ResultCode"] = result.code
	
	
	mdat = cbind(mdat, test.data)
	#print(c(length(test), length(names(mdat)), length(test.name)))
	names(mdat)[ncol(mdat)] = test.name

	if (trim.na) {
		mdat <- mdat[!is.na(test.data), ]
  }

	attributes(mdat)$Col.Meta <- as.list(local.col.meta)
	mdat
}

get.col.meta <- function(dataset) {
	if (!"Col.Meta" %in% names(attributes(dataset))) 
		stop("No column meta data found in the attributes. The attributes may have been stripped when the data.frame was modified.")
	col.meta = as.data.frame(attributes(dataset)$Col.Meta, stringsAsFactors=F)
  
	if (nrow(col.meta) != length(get.tests(dataset)))
	  stop("Number of tests doesn't match dimension of col.meta. This usually happens when you add a column to an mdat manually instead of using add.test.col()")
	row.names(col.meta) = get.tests(dataset)
	if ("TestType" %in% names(col.meta))
		col.meta$TestType = factor(col.meta$TestType, levels=test.type.levels)
	col.meta

}
```

```{r setup, echo=FALSE, message=FALSE}
#source("../scripts/functions/orig.R")
#source.packages()
library(dplyr)
library(tidyr)
library(magrittr)
library(DT)
library(knitr)
library(deming)
# This function loads all of our LIMS data. I'll replace it with fake data here:
#load.big.data()
fake.col.meta <- list(Name=c(), Abbrev=c(), UnitsOfMeasure=c(), ResultCode=c(), TestType=c(), Reported=c())
big.data <- data.frame(OriginalOrderDate <- sprintf("2016-%02i-01", sample(1:12, 1000, replace = T))) %>%
  {attributes(.)$Col.Meta <- fake.col.meta; .} %>%
  add.test.col(rnorm(1000, mean=200, sd=50), abbr="MPO", result.code='MPO', name="MPO", 
               units='pmol/L', reported = 1, type='Inflammation')
  
test.names <- get.test.names(big.data)

```


# Introduction
This short report summarizes the analysis of the new MPO assay. The analysis
includes comparing the old and new MPO assays and providing data to support new
reference range calculations.

# Data Preparation
Data for the current MPO assay was obtained from the Big Dataset, which is
current through August. The MPO data for the new MPO assay was provided by Hari
Nair. 

```{r data_prep}
# or
dats = list()
dats$orig <- big.data %>%
              filter(OriginalOrderDate <= "2016-09-01") %>% # Need this for future consistency 
              select.mdat(reported.only=T)
#dats$mpo <- read.csv("../datasets/CHL MPO Reference Range Data 09 16 16.csv")
dats$mpo <- data.frame(Result=rnorm(500, mean=220, sd=50))
#dats$side.by.side <- read.csv("../datasets/side_by_side.csv")
dats$side.by.side <- data.frame(Dimension=rnorm(500, mean=210, sd=50),
                                COBAS.Result=rnorm(500, mean=220, sd=50))
```

# Results

## Percentile Table
The following table compares quantiles between the current MPO assay and the
new MPO assay.

```{r section1}
old <- percentile.table(dats$orig[, test.names$MPO.code, drop=F])  %>%
  add_rownames("Percentile") %>%
  rename_("Current MPO"="MPO.codeMPO")

new <- percentile.table(dats$mpo[, "Result", drop=F]) %>%
  add_rownames("Percentile") %>%
  rename_("New MPO"="Result")

perc.table <- merge(old, new) %>%
  arrange(as.numeric(gsub("%", "", .$Percentile)))

kable(perc.table, digits=1)

```

## Quantile Concordance 

This plot shows the concordance between the percentiles listed in the previous
table. This plot reveals a bias toward larger values using the new MPO assay.

```{r concordance, fig.height=6, fig.width=6}
concord.plot(perc.table$Current, perc.table$New, xlab="Current MPO",
             ylab="New MPO", main="Quantile-Quantile Plot")
```

## Distribution Comparison

This plot shows the distributions of both MPO assays. This also reveals the new
MPO assay has slightly higher values. 

```{r fig.height=6, fig.width=6}
par(par.def)
list.pdf(list("Current MPO (Big Dataset)"=dats$orig$MPO.codeMPO,
              "New MPO (Ref Range Set)"=dats$mpo$Result),
         main="Distributions of MPO (Non-Paired)", xlab="MPO (pmol/L)", xlim=c(0, 500))

```

## Side by Side Comparison

This plot shows a comparison of samples run on the current and new MPO assays.
The x and y axes are shown on a log scale. The bulk of the distribution is
located on the concordance line. The new MPO assay signals higher for lower
valus and lower for high values. The two distributions are significantly
different at a 0.05 significance level by a Wilcoxon test and a log-transformed
t-test. The p-value was 0.02 for both tests. 
  
The percentile comparison plot has been overlayed on this plot in red. This
highlights that the samples used to calculate the new refernece range have
greater values than the side by side comparison. 

```{r, fig.height=6, fig.width=6}
concord.plot(dats$side.by.side$Dimension, dats$side.by.side$COBAS.Result,
             xlab="Current MPO (pmol/L)", ylab="New MPO (pmol/L)",
             main="Concordance between MPO Assays", log='', xlim=c(100, 750), ylim=c(100,750))
scatter.plot(perc.table$Current, perc.table$New, xlab="Current MPO",
             ylab="New MPO", main="Quantile-Quantile Plot", pch=19,
             col='red', add=T, include.cor = F, include.ab = T, line.col = 'red')
```

## Reference Range Estimation from Side by Side
  
```{r, fig.height=6, fig.width=6}
# Fake
ranges <- list(MPO.codeMPO=c(350, 633))
attributes(ranges$MPO.code604)$risks <- "OBH"
  
concord.plot(dats$side.by.side$Dimension, dats$side.by.side$COBAS.Result,
             xlab="Current MPO (pmol/L)", ylab="New MPO (pmol/L)",
             main="Concordance between MPO Assays", log='', xlim=c(0, 1000),
             ylim=c(0,1000), include.ab=F)

abline(47.65, .9, col='blue', lwd=2)
new.ranges <- ranges$MPO.codeMPO*.7 + 45
segments(-1000, new.ranges[1], ranges$MPO.codeMPO[1], col='red', lwd=2)
segments(-1000, new.ranges[2], ranges$MPO.codeMPO[2], col='red', lwd=2)
segments(ranges$MPO.codeMPO[1], -1000, y1=new.ranges[1], col='red', lwd=2)
segments(ranges$MPO.codeMPO[2], -1000, y1=new.ranges[2], col='red', lwd=2)
data.frame("Old Cutoff"=ranges$MPO.codeMPO, "New Cutoff"=new.ranges) %>% kable
```

## Current percentages

This table shows how many patients have Optimal, Borderline, and High Risk
designations on the current MPO assay. 

```{r section2, fig.width=4}
rr <- data.frame(MPO.codeMPO <- sample(c("O", "B", "H"), 1000, replace=T))

rr$MPO %>%
  perc.table %>%
  as.data.frame %>%
  {.$. %<>% replace.names(list(B="Borderline", H='High Risk', O="Optimal")); .} %>%
  {.$. %<>% factor(levels=c("Optimal", "Borderline", "High Risk")); .} %>% 
  rename_("Risk"='.', "Percentage"="Freq")  %>%
  arrange(Risk) %>%
  kable(digits=1)

```

## New ranges with same percentages

The following reference range would maintain the current percentages of optimal
(~70%), borderline (~25%), and high risk (~5%).

* Ref Cutoff 1: `r quantile(dats$mpo$Result, p=.7)`
* Ref Cuttof 2: `r quantile(dats$mpo$Result, p=.95)`

-----------------

# Discussion

The samples measured on the new MPO assay have greater MPO values that the big
dataset. The side by side comparison shows that values on the two assays don't
show the same bias observed between the reference range samples and the big
dataset. This may suggest that using these reference range samples will result
in reference ranges that are too high.


