---
title: "Lazyeval in base R"
author: Stu Field
date: "`r Sys.Date()`"
#output: rmarkdown::html_vignette
output:
  rmarkdown::html_document:
    fig_caption: yes
    code_folding: show
    number_sections: no
    toc: yes
    toc_float:
      collapsed: no
ratio: '9:16'
fontsize: 13pt
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
options(width = 100)
knitr::opts_chunk$set(
 collapse = TRUE,
 comment = "#>"
)
```

-------

### TL;DR

* promises special objects that are *un-evaluated* expressions
* use `substitute()` to access the promised expression, but not evaluate it
* use `eval()` to evaluate the expression in a specific environment
* data frames are lists & lists are environments
* *all* function arguments are initially promises


---------


#### Eval:

> "The expression given as function arguments are not evaluated
> before the function is called. Instead, the expressions are packaged
> together with the environment in which they should be evaluated and
> it is this package that is passed to the function.
> Evaluation only takes place when the argument is required."

#### Promises:

> "Promise objects are part of `R`’s lazy evaluation mechanism.
> They contain three slots: a value, an expression, and an environment.
> When a function is called the arguments are matched and then each of the
> formal arguments is bound to a promise. The expression that was given 
> for that formal argument and a pointer to the environment the function
> was called from are stored in the promise."



--------

#### Lazyeval

Lazy evaluation is a mechanism used by `R` (and many functional languages)
to delay the evaluation of an object until it is needed. To do this, `R`
has a data structure (object) called a "promise", which is an un-evaluated
expression, that can later be "fulfilled" by assigning it a value via
evaluation in a given environment.

Lazy evaluation (and "tidyeval") has and still remains a maddening, confusing,
and veritable Dante's Inferno of frustration for me. Every time I think
I have it nailed down, it turns around and slaps me in the face and says:
"oh no you don't!". It's a constant "2 steps forward, and 1.9 steps back".
This is an attempt to consolidate lazyeval gains and understanding, to
document what I currently know, and not have to start over every time
I need to dive into the lazyeval world yet again.


-------

### Examples of lazyeval

```{r ab}
f <- function(a, b) {
  # 'a' and 'b' are only promises, lazyeval means they are
  # only evaluated when asked for
  cat("Hello\n")
}
f()
```


```{r ifelse}
f <- function(a, b) {
  if ( TRUE ) {
    a
  } else {
    b
  }
}
f(4)
```

```{r ORpipe, error = TRUE}
f <- function(x) {
  if ( TRUE || x ) {   # x = promise that is never evaluated
    cat("Hello 1\n")
  }
  if ( TRUE | x ) {   # x = promise that IS evaluated -> Error
    cat("Hello 2\n")
  }
}
f()
```

```{r demo-time}
# note that each argument evaluated only when asked for
ping <- function(a = Sys.time(), b = Sys.time(), c = Sys.time()) {
  print(a)
  Sys.sleep(1)
  print(b)
  Sys.sleep(2)
  print(c)
  invisible()
}
ping()
```


### Leveraging lazyeval

> You can *intentionally* delay evaluation and 
  evaluate it "on demand" when YOU want.

Until that argument is accessed there is no value associated with
the promise. When the argument is accessed, the stored expression (symbol)
is evaluated in the stored environment, and the result is returned.
The result is also saved by the promise. The `substitute()` function will
extract the content of the expression slot. This allows the programmer to
access either the value OR the expression associated with the promise.

* `substitute()` captures a promise -> symbol
  - "expression slot extractor"
* `deparse()` converts expression/symbol -> character
* `eval()` evaluates the symbol in a given environment

True power of `substitute()` means you can capture a promise and evaluate
it at a time and in a context of *your* choosing.

```{r subst}
f <- function(x) {
  subx <- substitute(x)   # no error! keeps promise intact
  print(subx)
  chr <- deparse(subx)
  print(chr)
  invisible()
}
f(foo)

f(1:10)
```


Now we can understand what functions like the `dplyr` verbs `pull()`,
`mutate()` and `filter()` are doing under the hood:


```{r pull2}
# simplified `dplyr::pull()`
pull2 <- function(.data, var) {
  expr <- substitute(var)      # 1st capture the promise
  # remember:
  #   1) data frames are lists
  #   2) lists can behave like environments!
  eval(expr, envir = .data)   # 2nd evaluate it in context of `.data`
}

pull2(mtcars, mpg)
```

```{r mutate2}
# simplified `dplyr::mutate()`
mutate2 <- function(.data, expr_calc) {
  expr <- substitute(expr_calc)              # 1st capture the promise
  .data$new_var <- eval(expr, envir = .data) # 2nd evaluate it in context of `.data`
  .data
}

mutate2(head(mtcars), mpg * 2)
```

```{r mutate3}
# slightly more general `mutate2()`
mutate3 <- function(.data, ...) {
  dots <- match.call(expand.dots = FALSE)$... # similar to substitute(list(...))
  for ( var in names(dots) ) {
    .data[[var]] <- eval(dots[[var]], .data)
  }
  .data
}

mutate3(head(mtcars), mpg3x = mpg * 3, carb2 = carb ^ 2, vs_copy = vs)
```

```{r filter2}
# simplified `dplyr::filter()`
filter2 <- function(.data, ...) {
  lgl <- eval(substitute(...), .data)
  stopifnot(is.logical(lgl), length(lgl) == nrow(.data))
  .data[lgl, ]
}

filter2(mtcars, cyl == 8 & mpg > 15)
```



### Detecting promises

From the `R` manual:

> Within the R language, promise objects are almost only seen implicitly:
> actual function arguments are of this type. There is also a 
> `delayedAssign()` function that will make a promise out of an expression.
> There is generally
> **no way in R code to check whether an object is a promise or not,**
> **nor is there a way to use R code to determine the environment of a promise.**


So it appears you can *create* a promise, but not detect whether an object
is a promise or not. Boo! :(

### Forcing promises

To "force" a promise is nothing more than to assign a value to the promise
slot via evaluation. This can be accomplished either by calling the
promise variable, or by using the `force()` function.


```{r force, fig.height = 5, fig.width = 9}
par(mfrow = 1:2)
f <- function(x, y = deparse(substitute(x))) {
  x <- log(x)
  plot(x, main = y) # 'y' is forced AFTER 'x' has changed
  invisible()
}
f(1:10)

# we can force 'y' first
f2 <- function(x, y = deparse(substitute(x))) {
  force(y)
  x <- log(x)
  plot(x, main = y)
  invisible()
}
f2(1:10)
```

-------

### Resources

* [Roots of quotation](https://milesmcbain.xyz/posts/the-roots-of-quotation/)

* [r-bloggers: About lazy evaluation](https://www.r-bloggers.com/2018/07/about-lazy-evaluation/)
